---
title: "Weekly Exercise 6"
author: "YOUR_NAME"
output:
  html_document: 
    df_print: default
    theme: cerulean
    toc: yes
    code_folding: show
    toc_float: TRUE
    self_contained: true
editor_options: 
  chunk_output_type: console
---

# Notes

As now you already are familiarized with the basics of programming in R. 
From now on we are going to give you more general instructions to do the 
exercises, instead of step by step. So that you start to make your own decisions
on how to call your variables and how exactly you want to achieve the tasks.

For the following tasks, you will be asked to submit a .Rmd file, also known as
"R notebook". This code can be rendered by R into different publication 
formats. The upper part of this file is called "YAML header" and has some main 
options to render the notebook. The options in html_document: will tell R how 
to output the document as an HTML page (do not change that for submitting your 
exercises correctly).

At the end of the exercise you will find questions, for which you will need
to print the answers based on your code.

For submitting the exercise check the following guidelines:

1. Change the author option "YOUR_NAME" in the upper part of this file with your 
name.
2. Use only the code chunks to write your code, and the bottom part to write
your answers to the questions.
3. If you got slightly different plots to ours and don't manage 
to get the same plot as we do, don't frustrate and spend to much time to make 
it look identical. What we are most interested in is evaluating that you can do
the task and that your programming logic is sound (There is also a very small 
chance we mess things up at some point so don't attempt to make an identical
copy of our results ;) ).
4. We ask you to store the answer to some questions in the objects `ANSWERX`;
**use the code to store the answer, do not simply write the answer!**. If you do
this, we will consider it WRONG!
5. DO NOT leave in the code any unnecessary code such as View() and head()
that makes running and knitting the script clumsy.

```{r, Global Options, echo = F}
knitr::opts_chunk$set(warning = F, message = F, echo = T) # Don't change this
```

# 0. Load your packages

Load the packages you will use in this section.

```{r, Packages}
# YOUR CODE
```

# 1. Data wrangling (or fixing the data)

We are going to use a dataset from the United Kingdom government. 
The data set describes the average weekly intake of diverse products by region.
Our goal is to appreciate differences and pinpoint what are the products that
are the main drive for **differences between regions**.

(A) Load the tables "UK_weekFood_1.txt" and "UK_weekFood_2.txt" so that the first 
row in the file is the column names and that the first column is the rownames.
(B) Remove the units column in both datasets.
(C) Create a table that contains the data of all 13 cities, and shared food
categories that are in both datasets. You could use: `intersect()`, `match()`, `[`.
(E) Answer Question 1 using `setdiff()`, and store your answer in the object 
`ANSWER1`.

**Note: Understand what happens when you change the order of sets in the**
**setdiff function**

```{r, Data wrangling}
# YOUR CODE
ANSWER1 <- "YOURCODE"
```

# 2. Principal Components Analysis.

The full data with ingredients present in both files should be a dataframe
that has 19 food categories, and 13 regions. This essentially is explaining 13
data points by 19 dimensions. 

But we cannot easily visualize data in 19 dimensions, that is why we require 
mathematical tools like Principal Components Analysis (PCA) to transform the 
data and extract the more informative set of 'new dimensions', called Principal 
Components.

(A) To be able to run the PCA, first transpose the data so that food categories 
are the columns. (Important as PCA runs in the columns).
(B) Use the `prcomp()` function to perform a PCA on the data.
(C) Plot the first two principal components of the resulting new data object.
(D) Add a text label to each point by using the function `text()`, which works as
plotting dots in x and y coordinates but instead of dots writes words, and its 
arguments `cex` [numeric], `pos` [integer 1 to 4] and `col` [color code]. The
resulting plot does NOT need to look identical to the example plot, just make sure
that the labels of points make sense in the position they are.
(E) Answer question 2. Store your answer in the object ANSWER2.

```{r, Principal Components Analysis, fig.width= 6, fig.height=6}
# YOUR CODE
ANSWER2 <- "YOURCODE"
```

# 3. Looking for the most correlated (or anti correlated) variable

Now to understand were does most of the variability comes from we can look for
the most correlated variable to the first principal component. 

(A) Calculate the correlation between the PC1 and each of the variables. 
You can calculate this with the `cor()` function.
(B) Determine which is the most correlated (or anticorrelated variable). Remember
use functions to make your life easier. `cor()`, `abs()`, `which.max()`, `sort()`
are some examples of functions you can use to help you with this task. 
Store your answer to QUESTION3 in the object `ANSWER3`.
(C) Plot a scatterplot of the variable you found as a function of the PC1.

Notes: Instead of "Mistery Food" as ylab, write the name of the variable 
you found. If you want to adjust the `blue` color use `adjustcolor()` is optional.

```{r, Most correlated variable, fig.width=6, fig.height=6}
# YOUR CODE
ANSWER3 <- "YOURCODE"
```

(D) OPTIONAL: Try to make the same scatterplot using ggplot. If you run into
trouble try to follow this 
[Quick guide](http://www.sthda.com/english/wiki/ggplot2-scatter-plots-quick-start-guide-r-software-and-data-visualization)
to troubleshoot your issues. Additionally you can adjust the range of the axis
using the `scale_x_continuous()` and `scale_y_continuous()` functions, this to
show the full labels (optional).

```{r, with ggplot, fig.width=6, fig.height=6}
# YOUR CODE
```

# Questions

By previously storing your answers on the `ANSWER1`, `ANSWER2`, `ANSWER3` objects
Rmarkdown will subtitute the lines bellow with the content of the object in 
the document.

* Q1: Which are the missing food categories? 

A: **`r ANSWER1`**

* Q2: How much variance does the PC1 explain?

A: **`r ANSWER2`**

* Q3: Which food category is the most correlated to PC1?

A: **`r ANSWER3`**


# Credits

Authors:

* Benjamin Dubreuil
* [Miguel Angel Garcia-Campos](https://angelcampos.github.io/)
* Yaara Finkel
* Hugo Schweke


Original data from:

* UK's government site: https://www.gov.uk/government/statistical-data-sets/family-food-datasets
* Direct link: https://assets.publishing.service.gov.uk/government/uploads/system/uploads/attachment_data/file/830364/ConsGORHH_12sep19.ods
